Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-04-26 SIR_manuscript.tex
discussion

Mathematical modeling of infectious disease outbreaks provides insights on how testing processes influence the epidemiological processes through isolation. 
Here, we develop a compartmental SIR-type model to study the potential effect of testing strategies, testing intensity, test sensitivity and specificity, test reporting time and isolation on epidemic dynamics. 
While targeted testing strategies 
\david{you need to be clearer: you mean targeting individuals who have contacted people who were infected.  For example, here you could say something like ``While targeting the contacts of confirmed cases\dots''}
are always more effective than random testing, as expected, we find that in some cases the direct effect of testing is that viral spread
is greater for a slow test than for a fast test. This counter-intuitive effect can occur when people are cautious when awaiting a test result, and may not be robust to second-order effects of fast testing (such as better contact tracing).

We incorporated the compartment-specific relative testing weights, $w_S$, $w_I$ and $w_R$, to model random testing and \emph{targeted} testing strategies. Here, in the case of targeted testing and for the simplicity and illustration purposes, we assumed that infected and recovered individuals are tested at three times the \percap rate of susceptible individuals, thus $w_I/w_S=3$ and $w_R=w_I$. Note that we have not specified a methodology to assign particular relative testing weights corresponding to a particular targeted testing scenario. 
This part needs to be developed further in future work. 
Modeling different targeted testing strategies, equivalently test-specific testing weights in our framework, requires prior information of the conditional probabilities of getting tested for people in a given compartment. 
This can be implied when we would like to quantify and compare the effect of different levels of test focus for infectious people on the basic reproduction number $\Rnum$, and conclude about the disease spread management. For example, when people are tested for "screening", the individuals with potential higher mobility, eg. people who are getting on flights, get more tested and thus the coresponding heavier testing weight is assigned than people awaiting for a surgery and are probably going to stay in a long-term care facility and consequently less mobile and more isolated to begin with. With our model, we would be able to compare the sensitivity of the disease dynamics, through $\Rnum$, with respect to testing processes.

{\bf The \percap testing intensity, $\rho$;}
Proposition \ref{prop1} part \ref{p1:rho} and also Figure \ref{pan} indicates that increasing testing intensity $\rho$ reduces $\Rnum$. This is sensible since as people are moved to test compartments, namely $X_p$ and $X_n$ for $X \in \{S,I,R\}$, they may partially or fully self-isolate. Also individuals who have tested positive, namely $X_c$ compartment for $X \in \{S,I,R\}$, are highly likely to self-isolate. The higher probability of being subject to isolation, the lower the force of infection \eqref{Lambda} and the lower $\Rnum$ will become. While our simulation supported this result, analytically it appeared to be hard to conclude this due to the complexity of $\Rnum$ expression.

>>>>

Mathematical modeling of infectious disease outbreaks provides insights on how testing processes influence the epidemiological processes through isolation. 
Here, we develop a compartmental SIR-type model to study the potential effect of testing strategies, testing intensity, test sensitivity and specificity, test reporting time and isolation on epidemic dynamics. 
While targeted testing strategies, including targeting people with infection-like symptoms or the contacts of confirmed cases, are always more effective than random testing, as expected, we find that in some cases the direct effect of testing is that viral spread is greater for a slow test than for a fast test. This counter-intuitive effect can occur when people are cautious when awaiting a test result, and may not be robust to second-order effects of fast testing (such as better contact tracing). \ali{this is a valid sentence especially when $\theta\_w$ is closer to 1, but not sure if this was counter-intuitive or was the opposite of it? i.e., that viral spread is greater for a fast test ($1/\omega$ closer to 0) than for a slow test.}

We incorporated the compartment-specific relative testing weights, $w_S$, $w_I$ and $w_R$, to model random testing and \emph{targeted} testing strategies. Here, in the case of targeted testing and for the simplicity and illustration purposes, we assumed that infected and recovered individuals are tested at three times the \percap rate of susceptible individuals, thus $w_I/w_S=3$ and $w_R=w_I$. Note that we have not specified a methodology to assign particular relative testing weights corresponding to a particular targeted testing scenario. 
Modeling different targeted testing strategies, equivalently test-specific testing weights in our framework, requires prior information of the conditional probabilities of getting tested for people in a given compartment. 
This can be implied when we would like to quantify and compare the effect of different levels of test focus for infectious people on the basic reproduction number $\Rnum$, and conclude about the disease spread management. For example, when people are tested for "screening", the individuals with potential higher mobility, eg. people who are getting on flights, get more tested and thus the coresponding heavier testing weight is assigned than people awaiting for a surgery and are probably going to stay in a long-term-care facility and consequently less mobile and more isolated to begin with. With our model, we would be able to compare the sensitivity of the disease dynamics, through $\Rnum$, with respect to testing different high-risk groups in the population. This part needs to be developed further in future work.

{\bf The \percap testing intensity, $\rho$;}
Proposition \ref{p1:rho} and also Figures \ref{pan} and \ref{pan2} indicates that increasing testing intensity $\rho$ may reduce $\Rnum$ but there is a counter-intuitive machanism. 
The expected mechanism was that as more untested infected individuals are moved to the test compartments, i.e., from $I\_u$ to $I\_p$ and $I\_n$, and eventually to the confirmed compartment $I\_c$, they may partially or fully self-isolate ($\theta\_w$ and $\theta\_c$ varies between 0 and 1 with $\theta\_w\leq\theta\_c$). Thus, the higher the \percap testing intensity $\rho$, the lower the force of infection \eqref{Lambda} and the lower $\Rnum$ will become. This is true when $\rho$ is very small relative to $N_0$ and can be verified analytically and through our simulation. The counter-intuitive mechanism is that at the DFE, higher \percap testing intensity, may delay recognition of true infected individuals $I\_p$ and causing an increase in $\Rnum$ as follows. At high $\rho$, more people will be moved from $S\_u^*$ to $S\_n^*$, thus a higher proportion of awaiting people may get infected and end up in $I\_n$, and be delayed from being eligable for retesting $I\_u$ and eventually to $I\_p$ and $I\_c$. Our analysis (Sec.\,\ref{app:rho}) and simulation (see Fig.\,\ref{pan2} panel (b) when $\theta\_c=1$ and $\theta\_w=0$) support such a mechanism. In particular, if \eqref{ineq:a3} holds we will have such a result.
